\bf Model Visualization Techniques for a Social Network Model

### 2018-05-29 paper-child.Rnw
it compiles now! :smile:

We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figures \ref{fig:adjmatviz, fig:alldataviz}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align*}

@@

numordersf$from <- factor(numordersf$from)
levels(numordersf$from) <-  1:16
numordersf$to <- factor(numordersf$to)
levels(numordersf$to) <- 1:16
numordersf$from <- as.numeric(as.character(numordersf$from))
numordersf$to <- as.numeric(as.character(numordersf$to))

p1 <- ggplot(data = numordersf, aes(y = -from, x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_continuous(breaks = 1:16) +
  scale_y_continuous(breaks = (1:16)*-1, labels =1:16)  +
  ThemeNoNet +
  theme(legend.position = 'none', panel.grid = element_blank(), 
        panel.border = element_blank(), 
        panel.spacing = unit(rep(0,4),"cm")) +
  coord_fixed()

@@

pca1 <- prcomp(allmsrep1_adjmat)
order1 <- names(sort(pca1$rotation[,1], decreasing = T))
order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))

pcaordersf$from <- factor(pcaordersf$from, levels = rev(order1))
pcaordersf$to <- factor(pcaordersf$to, levels = order1)

p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top", labels = str_replace(order1, "V", "")) +
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet +
  theme(legend.position = 'none') +
  coord_fixed()

@@

<<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes. The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
ansnullchains %>% group_by(period, rep) %>%
  mutate(ms = row_number()) %>%
  filter(ms == 1, period == 1) %>%
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>%
  ungroup() %>%
  group_by(ego, alter) %>%
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>%
  group_by(ego) %>%
  mutate(csum = cumsum(alterProb)) %>%
  arrange(ego, csum) %>%
  mutate(totego = sum(count),
         empprob = count / totego) -> cumprobs
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique

wave1friends2 <- wave1friends[,-3]
wave1friends2$type <- "remove"
changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))

@@

<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations. The acting ego node is on the y-axis, and the alter node is on the x-axis. There were many ties with empirical probability zero.', fig.height=2.5>>=
ansnullchains %>%
  filter(period == 1) %>%
  group_by(rep) %>%
  mutate(msno = row_number()) %>%
  filter(msno == 1) %>%
  mutate(ego = paste0("V",as.numeric(as.character(ego))+1),
         alter = paste0("V",as.numeric(as.character(alter))+1),
         prob = exp(logAlterProb)) -> probsDat

# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
# 103, 3
ggplot(data = probsDat,
       aes(y = factor(ego, levels = order1),
             x = factor(alter, levels = rev(order1)),
             fill = prob)) +
  geom_tile(color = 'black') +
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  labs(y = "ego", x = "alter")+
  scale_x_discrete(position = 'top', labels = str_replace(order1, "V", "")) +
  scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + theme(panel.grid = element_blank()) +
  coord_fixed()
@

@@

<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5>>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to),
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>%
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>%
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\ref{sec:allpossible}. Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.">>=
smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
library(ggrepel)
set.seed(12345678)
smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
ggplot(aes(x = estimate, y = Waldpval, color = inter1)) +
  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = shortName)) +
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for friendship data") +
  coord_flip() +
  ThemeNoNet
sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) +
  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = shortName)) +
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for senate data") +
  coord_flip() +
  ThemeNoNet
@

@@

adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
adjmatall2 <- data.frame(adjmatall2)
adjmatall2$from <- rep(paste0("V", 1:16), 56)
names(adjmatall2)[17] <- "ms"
adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3

adjmatall3$to <- factor(adjmatall3$to, levels = order1)
adjmatall3$from <- factor(adjmatall3$from, levels = rev(order1))

>>>>

We first consider the 16 actor subset of the teenage friends and lifestyle data show in Figure~\ref{fig:alldataviz}. To this data, we fit three different SAOMs. Each SAOM used a simple rate function, $\alpha_m$, and an objective function with two or three parameters. The first model, M1, contains the absolute minimum number of parameters in the objective function $f_i(x)$:  $$f_i(x)^{M1} = \beta_1s_{i1} + \beta_2s_{i2},$$ where $s_{i1}$ is the density network statistic and $s_{i2}$ is the reciprocity network statistic for actor $i$ at the current network state $x$. The second and third models, M2 and M3, contain one additional parameter each in the objective function which were determined by a Wald-type test provided in the \texttt{RSiena} software to be significant, with $p$-values less than 0.05 \citep{RSienaTest}. The M2 model contains an actor-level covariate parameter, and the M3 model contains an additional strutural effect in the objective function.
\begin{align*}
f_i(x)^{M2} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_3s_{i3} \\
f_i(x)^{M3} & = \beta_1s_{i1} + \beta_2s_{i2} + \beta_4s_{i4},
\end{align*}

@@

numordersf$from <- factor(numordersf$from)
levels(numordersf$from) <-  1:16
numordersf$to <- factor(numordersf$to)
levels(numordersf$to) <- 1:16

p1 <- ggplot(data = numordersf, aes(y = rev(from), x = to)) + geom_tile(color = 'black', aes(fill = as.factor(value))) +
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(labels = 1:16, position = "top") +
  scale_y_discrete(labels =16:1, name = "from")  +
  ThemeNoNet +
  theme(legend.position = 'none', panel.grid = element_blank(), 
        panel.border = element_blank(), 
        axis.text.y = element_text(margin = margin(r=0))) +
  coord_fixed()

@@

pca1 <- prcomp(allmsrep1_adjmat)
order1 <- names(sort(pca1$rotation[,1], decreasing = T))
order1 <- gsub("V", "",order1)
# order2 <- names(sort(abs(pca1$rotation[,1]), decreasing = T))

pcaordersf$from <- as.factor(pcaordersf$from)
levels(pcaordersf$from) <- 1:16
pcaordersf$from <- factor(pcaordersf$from , levels = levels(pcaordersf$from)[match( order1,levels(pcaordersf$from) )])
pcaordersf$to <- as.factor(pcaordersf$to)
levels(pcaordersf$to) <- 1:16
pcaordersf$to <- factor(pcaordersf$to , levels = levels(pcaordersf$to)[match( order1,levels(pcaordersf$to) )])

p2 <- ggplot(data = pcaordersf, aes(x = to, y = from)) +
  geom_tile(color = 'black', aes(fill = as.factor(value))) +
  scale_fill_manual(values = c('white', 'black')) +
  scale_x_discrete(position = "top") +
  #scale_y_discrete(name = "from",labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet +
  theme(legend.position = 'none') +
  coord_fixed()

@@

<<histogramalters, fig.width = 7, fig.height=9, fig.cap="Each panel shows the theoretical (as lines) and empirical (as points) probabilities of the chosen ego node changing its tie to each of the other nodes. The color of the line indicates whether the tie from the ego to the alter node is being added, removed, or if there is no change to the network in this step.">>=
set.seed(823746)
load("data/ansnullpaper.rda")
ansnullchains <- get_chain_info(ansnull)
ansnullchains %>% group_by(period, rep) %>%
  mutate(ms = row_number()) %>%
  filter(ms == 1, period == 1) %>%
  mutate(alterProb = exp(logAlterProb),
         ego = as.numeric(as.character(ego)),
         alter = as.numeric(as.character(alter))) %>%
  ungroup() %>%
  group_by(ego, alter) %>%
  summarize(count = n(), alterProb = unique(alterProb)) %>% ungroup() %>%
  group_by(ego) %>%
  mutate(csum = cumsum(alterProb)) %>%
  arrange(ego, csum) %>%
  mutate(totego = sum(count),
         empprob = count / totego) -> cumprobs
cumprobs$ego <- factor(paste0("V",cumprobs$ego+1), levels = paste0("V", 1:16))
cumprobs$alter <- factor(paste0("V",cumprobs$alter+1), levels = paste0("V", 1:16))

@@

changes <- cumprobs %>% dplyr::select(ego, alter) %>% unique
wave1friends <- fortify(as.adjmat(fd2.w1))
wave1friends2 <- wave1friends[,-3]
wave1friends2$type <- "remove"
changes2 <- left_join(changes, wave1friends2, c("ego" = "from", "alter" = "to"))

@@

<<heatmap, fig.align='center', fig.cap='A heatmap showing the empirical transition probabilities for the first microstep in 1,000 simulations. The acting ego node is on the y-axis, and the alter node is on the x-axis. There were many ties with empirical probability zero.', fig.height=2.5>>=
ansnullchains %>%
  filter(period == 1) %>%
  group_by(rep) %>%
  mutate(msno = row_number()) %>%
  filter(msno == 1) %>%
  mutate(ego = as.numeric(as.character(ego))+1,
         alter = as.numeric(as.character(alter))+1,
         prob = exp(logAlterProb)) -> probsDat

# probsDat %>% group_by(ego, alter) %>% summarize(count = n()) %>% dim
# 103, 3
ggplot(data = probsDat,
       aes(y = factor(ego, levels = order1),
             x = factor(alter, levels = order1),
             fill = prob)) +
  geom_tile(color = 'black') +
  scale_fill_gradient(low = 'white', high = 'blue', limits = c(0,1)) +
  labs(y = "ego", x = "alter")+
  scale_x_discrete(position = 'top') +
  #scale_y_discrete(labels = rev(str_replace(order1, "V", ""))) +
  ThemeNoNet + 
  theme(panel.grid = element_blank()) +
  coord_fixed()
@

@@

<<allsteps, fig.cap="Two simulations (out of 1,000) of the microstep process from wave 1 to wave 2. The $x$ axis is the microstep number, with step 0 representing the first wave of data and the final step representing the second wave of data. We can see that many edges are underrepresented in this process: they are in the second wave, but never appear in the microsteps.", fig.height=2.5>>=
smallfriends <- read_csv("data/smallfriends4Geomnet.csv")
names(smallfriends)[1:2] <- c("from", "to")
ansnullchains %>%
  dplyr::filter(period == 1) %>%  #only look at chains from wave 1 to wave 2
  #dplyr::group_by(rep) %>%
  dplyr::select(rep, from = ego, to = alter) %>%
  dplyr::mutate(val = as.numeric(!from == to),
         from = paste0("V", parse_number(from)+1), # make the chains
         to = paste0("V", parse_number(to)+1)) -> ansnullchainsw1w2
msall1 <- netvizinf::listMicrosteps(dat = wave1friends,
                      microsteps = filter(ansnullchainsw1w2, rep == 1))
msall1_df <- plyr::rbind.fill(lapply(X = seq_along(msall1), FUN = function(i, x) {
        x[[i]]$ms <- i - 1
        return(x[[i]])
    }, msall1))
msall1_df <- msall1_df %>% dplyr::select(from, to, ms)
wave2 <- smallfriends %>% filter(Wave == 2) %>% na.omit %>% data.frame()
msall1_df <- msall1_df %>% add_row(from = paste0("V",wave2$from),
                      to = paste0("V",wave2$to),
                      ms = max(msall1_df$ms)+1)
edges <- msall1_df %>%
  group_by(from, to) %>%
  summarize(edgeappears = n()) %>%
  arrange(desc(edgeappears)) %>% ungroup() %>%
  mutate(eid = row_number())

@@

<<sigtesting, fig.show = 'hold', out.width='\\textwidth', fig.height=2, fig.cap="From Section~\ref{sec:allpossible}. Significant effects for the two data sets, at a significance level of 0.10 or lower as calculated by the Wald-type test available in the SIENA software.", error=FALSE>>=
smfr_effs <- read_csv("data/effects_significance_smallFriends.csv")
smfr_effs$inter1[is.na(smfr_effs$inter1)] <- "none"
library(ggrepel)
set.seed(12345678)
smfr_effs %>% filter(type == "eval" & Waldpval <=0.10) %>%
ggplot(aes(x = estimate, y = Waldpval, color = inter1)) +
  #geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = shortName)) +
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for friendship data") +
  coord_flip() +
  ThemeNoNet
sen_effs <- read_csv("data/congress/senateModelSigEffects.csv")
#sen_effs$Waldpval <- parse_number(sen_effs$Waldpval)
sen_effs$inter1[is.na(sen_effs$inter1)] <- "none"
ggplot(data = sen_effs %>% filter(Waldpval <=0.10), aes(x = estimate, y = Waldpval, color = inter1)) +
  # geom_errorbarh(aes(xmin = estimate - se, xmax = estimate + se), color = 'black') +
  geom_vline(xintercept = 0) +
  geom_point(size = 3) +
  #geom_text_repel(aes(label = shortName)) +
  scale_color_brewer(palette = "Paired", name = "Interaction\nVariable") +
  ggtitle("Significant effects at level 0.10 for senate data") +
  coord_flip() +
  ThemeNoNet
@

@@

adjmatall2 <- cbind(adjmatall2, rep(0:55, each = 16))
rownames(adjmatall2) <- NULL
adjmatall2 <- as.data.frame(adjmatall2)
adjmatall2$from <- rep(paste0("V", 1:16), 56)
names(adjmatall2)[17] <- "ms"
adjmatall2 %>% gather(to, val, V1:V16) -> adjmatall3

adjmatall3$to <- as.factor(adjmatall3$to)
levels(adjmatall3$to) <- gsub("V", "", levels(adjmatall3$to))
adjmatall3$from <- as.factor(adjmatall3$from)
levels(adjmatall3$from) <- gsub("V", "", levels(adjmatall3$from))
adjmatall3$to <- factor(adjmatall3$to, levels = order1)
adjmatall3$from <- factor(adjmatall3$from, levels = order1)
